Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Erythranthe cardinalis}

### 2017-09-02 ms/ms.Rnw
effect sizes calculations

  # Source custom functions
  #source(paste(pathMS, "/functions.R", sep = ""))
  source("functions.R")

@@

% Intrinsic versus plastic variation 
When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004}. The traits that underlie local adaptation, however, need not mirror this pattern. Populations can have fixed genetic differences conferring trait values that are adaptive at home but neutral or maladaptive away. Alternatively, the ability to plastically respond to a particular environment or the magnitude of response to an environment could be adaptive. We distinguish between these patterns of adaptive trait differences by referring to `intrinsic' and `plastic' trait variation, respectively. Both intrinsic and plastic trait variation can be explained by genetic differences and both are involved in adaptation. For example, intrinsic differences in photoperiod responses \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment. Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}. In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.

% Selective agents (trait-enviro assoc)
A basic approach to identify candidate traits underlying local adaptation is to find associations between traits and environments. Either intrinsic and/or plastic variation should vary clinally along environmental gradients. Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown. For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}. Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation. Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, leaf morphology \citep{Hopkins_etal_2008, Stock_etal_2014}, and drought response \citep{Kooyers_etal_2015} that likely relate to climatic variation.

@@

In this study, we investigated two major questions: 1) whether intrinsic or plastic physiological trait variation corresponds with latitude; and 2) what climatic factor(s) could plausibly be responsible for latitudinal clines. Within question 2, we tested three hypotheses outlined in the previous paragraph: latitudinal clines are explained by a single dominant climatic factor, multiple climatic factors, or the climatic neighborhood experienced by nearby population connected through gene flow. These hypotheses are not mutually exclusive since, for example, single or multiple factors in a climatic neighborhood may lead to latitudinal clines. We focused on climate because climate often determines and where species are found and also can exert strong selection on populations within species, though we acknowledge that other abiotic and biotic factors could also contribute to selection and the overall pattern of local adaptation. There is also a compelling need to know how populations are (or are not) locally adapted to climate so as to predict how they will respond to climate change \citep{Aitken_Whitlock_2013}.

We examined these questions in \textit{Erythranthe cardinalis} (formerly \textit{Mimulus cardinalis} [\citeauthor{Nesom_2014} \citeyear{Nesom_2014}]) because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, and genomic studies under both lab and field conditions. Many classic and contemporary studies of local adaptation use \textit{Mimulus sensu lato} species because of their natural history, easy propagation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}. Yet, there is a deficiency of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010, Wright_etal_2013}. We measured genetic and genotype-by-environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude. We found a latitudinal cline of intrinsic variation in photosynthesis and growth, but little evidence for variation in plasticity. Interannual variation in precipitation and temperature are associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{E. cardinalis}. The climatic neighborhoods around populations explained trait variation better than local climate, indicating that latitudinal clines may be common because latitude integrates effects of selection on populations connected through gene flow. We place these findings in the context of life history theory and consider future directions in the Discussion.

@@

\textit{E. cardinalis} is a perennial forb native to the Western US (California and Oregon). It is predomintantly outcrossing, self-compatible, and pollinated primarily by hummingbirds. We used 16 populations from throughout the range of \textit{E. cardinalis} (Table~\ref{table:Table_FocalPops}). These populations were intentionally chosen to span much of the climatic range of the species based on all known occurrences (see below). Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature. We used seeds from \Sexpr{nlevels(master$Family)} families, \Sexpr{min(famTab)}--\Sexpr{max(famTab)} (mean = \Sexpr{round(mean(famTab), 1)}, median = \Sexpr{median(famTab)}) families per population.

@@

# Rug indicating measurment timing
rect(6, -2, 17, 0, border = "white", col = "grey")
with(exptlDes, text(11.5, -1, labels = "Germ", col = "white", font = 3, 
  cex = 0.9))

with(exptlDes, rect(min(which(LERmeas)), -2, max(which(LERmeas)), 0, 
	border = "white", col = "grey"))
with(exptlDes, text(min((which(LERmeas)) + max(which(LERmeas))) / 2, -1, 
	labels = "LER", col = "white", font = 3, cex = 0.9))

with(exptlDes, rect(min(which(SERmeas)), 0, max(which(SERmeas)), 2,
	border = "white", col = "grey"))
with(exptlDes, text(min((which(SERmeas)) + max(which(SERmeas))) / 2 - 5, 1, 
	labels = "SER", col = "white", font = 3, cex = 0.9))

with(exptlDes, rect(min(which(Harvest)), -2, max(which(Harvest)), 0, 
	border = "white", col = "grey"))
with(exptlDes, text(min((which(Harvest)) + max(which(Harvest))) / 2, -1, 
	labels = "Harvest", col = "white", font = 3, cex = 0.9))

with(exptlDes, text(min((which(PhotoMeas)) + max(which(PhotoMeas))) / 2, 5, 
	labels = "Photo", col = "grey", font = 3, cex = 0.9))
with(exptlDes, rect(min(which(PhotoMeas)), 2, max(which(PhotoMeas)), 4, 
	border = "white", col = "grey"))

@@

@@

@@

\paragraph{Photosynthesis}
During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate on a subset of \Sexpr{length(which(!is.na(master$Photo)))} plants evenly spread between treatments and families within populations. The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6-cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska). We made all measurements at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$ of photosynthetically active radiation), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity. During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to the average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$) so that all plants within a temperature level could be measured under the same conditions.

@@

@@

  # Figure of reaction norms
  d <- subset(master, !is.na(master$DiedFromStress))
  tab <- tapply(d$DiedFromStress, list(d$PopID, d$WaterTrt, d$TempTrt), 
                function(X) 1 - length(which(X)) / length(X))
  pdf(paste(pathFig, "/FigureS_MortRxnNorm.pdf", sep = ""), 4, 4, useDingbats = F)
  par(mai = c(0.5, 1, 0.5, 0.25))
  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, 
       xlim = c(1, 4), ylim = c(0, 1))
  abline(v = 2.5)
  mtext(side = 3, c("Cool", "Hot"), at = c(1.625, 3.375))
  mtext(side = 1, c("Wet", "Dry", "Wet", "Dry"), at = c(1, 2.25, 2.75, 4))
  mtext(side = 1, "Water Treatment", font = 2, cex = 1.25, line = 1.5)
  mtext(side = 3, "Temperature Treatment", font = 2, cex = 1.25, line = 1.5)
  axis(2, lwd = 0, lwd.ticks = 1, las = 1)
  title(ylab = "Survivorship", cex.lab = 1.5)

@@

\subsection*{Intrinsic variation and plasticity}

For all traits (Table~\ref{table:Table_traits}) except germination (see above), we tested for Population, Treatment (Temperature, Water, and Temperature $\times$ Water), and Population $\times$ Treatment interactions (Population $\times$ Temperature, Population $\times$ Water, and Population $\times$ Temperature $\times$ Water). We interpreted significant Population effects to indicate intrinsic variation and Population $\times$ Treatment interactions to indicate variation in plasticity. As mentioned above, we used survival and GLMM models for germination rate and mortality, respectively. For all other traits, we used mixed model ANOVAs with Family and Block included as random factors. We fit models using restricted maximum likelihood in lmer, a function in the R package \pkg{lme4} \citep{Bates_etal_2015}. We determined significant fixed effect terms using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-32 \citep{Kuznetsova_etal_2016}. This package uses Satterthwaite's approximation to calculate denominator degrees of freedom for $F$-tests. We also included days to germination as a covariate in growth analyses. To ensure that Population and Treatment effects were specific to a particular growth phase, we included germination day as a covariate in leaf expansion and stem elongation analyses.

@@

We selected Climate-Latitude and Climate-TraitPC1 variables independently using Variable Selection Using Random Forest (VSURF) algorithm in the R package \pkg{VSURF} version 1.0.3 \citep{Genuer_etal_2016}. Random Forest regression is useful for cases like ours when the number of potential predictors is similar to or greater than the number of observations (`high $p$, low $n$' problem). VSURF is a multistip algorithm that progressively retains or eliminates variables based on their importance over regression trees in the forest. Variable importance is defined as the average amount a climate variable reduces mean-squared error in the predicted response (TraitPC1 or Latitude), compared to a randomly permuted dataset, across all trees in the random forest (see \citeauthor{Genuer_etal_2015} [\citeyear{Genuer_etal_2015}] for further detail). Hence, VSURF automatically eliminates unimportant and redundant variables based on the data without having to arbitrarily choose among colinear climate variables before the analysis. We kept only variables selected for prediction, the most stringent criterion. A visual overview of how we selected climatic variables is depicted in Fig~\ref{fig:Fig_VarSelect}.

<<Fig_VarSelect, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=

@@

Genotype $\times$ environment (G$\times$E) interactions are also a common signature of local adaptation. In contrast to the intrinsic differences described above, we found little evidence of G$\times$E in \textit{E. cardinalis}. There were only two statistically significant Population $\times$ Treatment interactions (Table~\ref{table:Table_AnovaSummary}, Fig.~\ref{fig:FigS_RxnNorms}), but these were not strong compared to Population and Temperature effects. Otherwise, populations responded similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}). Complete ANOVA tables are available in the Supporting Information (\Cref{table:TableS_fitLER,table:TableS_fitSER,table:TableS_fitPhoto,table:TableS_fitMort})

@@

Interannual variation in climate averaged over each populations's climatic neighborhood correlated most strongly with trait variation and latitude of \textit{E. cardinalis} occurrences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}). All \Sexpr{nImpLat} Climate-Latitude and \Sexpr{nImpPC1} Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}). In fact, neighborhood climate almost always correlated better with TraitPC1 and Latitude than local climate (Fig.~\ref{fig:Fig_LocalvNeighCorr}). On average, neighborhood Climate-TraitPC1 correlation coefficients were \Sexpr{-round(tt1$estimate, 2)} higher than correlations with local-scale climate variables (paired \textit{t}-test, $t = \Sexpr{-round(tt1$statistic, 2)}$, d.f. = \Sexpr{round(Meff1 / 2 - 1, 1)}, $P = \Sexpr{round(pt1 * 10 ^ -floor(log10(pt1)), 2)} \times 10 ^ {\Sexpr{floor(log10(pt1))}}$). Likewise, neighborhood Climate-Latitude correlation coefficients were \Sexpr{-round(tt2$estimate, 2)} higher than those for local-scale climate (paired \textit{t}-test, $t = \Sexpr{-round(tt2$statistic, 2)}$, d.f. = \Sexpr{round(Meff2 / 2 - 1, 1)}, $P = \Sexpr{round(pt2 * 10 ^ -floor(log10(pt2)), 2)} \times 10 ^ {\Sexpr{floor(log10(pt2))}}$).

>>>>

  # Source custom functions
  source("functions.R") # for sweave
  # source(paste(pathMS, "/functions.R", sep = "")) # for debugging

@@

% Intrinsic versus plastic variation 
When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004}. The traits that underlie local adaptation, however, need not mirror this pattern. Populations can have fixed genetic differences conferring trait values that are adaptive at home but neutral or maladaptive away. Alternatively, the ability to plastically respond to a particular environment or the magnitude of response to an environment could be adaptive. We distinguish between these patterns of adaptive trait differences by referring to `genetic variation' in trait means and `genetic variation in plasticity', respectively. Genetic variation in plasticity is synonymous with genotype-by-environment interactions, or simply (G$\times$E). Genetic variation in trait means and plasticity are both involved in adaptation. For example, genetic variation in photoperiod responses \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment. Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genetic variation in plasticity cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}. In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance \citep{Flood_etal_2011}.

% Selective agents (trait-enviro assoc)
A basic approach to identify candidate traits underlying local adaptation is to find associations between traits and environments. Either genetic variation in trait means and/or plasticity should vary clinally along environmental gradients. Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown. For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}. Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation. Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, leaf morphology \citep{Hopkins_etal_2008, Stock_etal_2014}, and drought resistance \citep{Kooyers_etal_2015} that likely relate to climatic variation.

@@

In this study, we investigated two major questions: 1) whether genetic variation in physiological trait means or plasticity corresponds with latitude; and 2) what climatic factor(s) could plausibly be responsible for latitudinal clines. Within question 2, we tested three hypotheses outlined in the previous paragraph: latitudinal clines are explained by a single dominant climatic factor, multiple climatic factors, or the climatic neighborhood experienced by nearby population connected through gene flow. These hypotheses are not mutually exclusive since, for example, single or multiple factors in a climatic neighborhood may lead to latitudinal clines. We focused on climate because climate often determines and where species are found and also can exert strong selection on populations within species, though we acknowledge that other abiotic and biotic factors could also contribute to selection and the overall pattern of local adaptation. There is also a compelling need to know how populations are (or are not) locally adapted to climate so as to predict how they will respond to climate change \citep{Aitken_Whitlock_2013}.

We examined these questions in \textit{Erythranthe cardinalis} (formerly \textit{Mimulus cardinalis} [\citeauthor{Nesom_2014} \citeyear{Nesom_2014}]) because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, and genomic studies under both lab and field conditions. Many classic and contemporary studies of local adaptation use \textit{Mimulus sensu lato} species because of their natural history, easy propagation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}. Yet, there is a deficiency of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010, Wright_etal_2013}. We measured genetic variation in trait means and plasticity in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude. We found a latitudinal cline of trait means in photosynthesis and growth, but little evidence for variation in plasticity. Interannual variation in precipitation and temperature are associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{E. cardinalis}. The climatic neighborhoods around populations explained trait variation better than local climate, indicating that latitudinal clines may be common because latitude integrates effects of selection on populations connected through gene flow. We place these findings in the context of life history theory and consider future directions in the Discussion.

@@

\textit{E. cardinalis} is a perennial forb native to the Western US (California and Oregon). It is predomintantly outcrossing, self-compatible, and pollinated primarily by hummingbirds. We used 16 populations from throughout the range of \textit{E. cardinalis} (Table~\ref{table:Table_FocalPops}). These populations were intentionally chosen to span much of the climatic range of the species based on all known occurrences (see below). Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature. To control for maternal effects, we grew a large number of field-derived seeds in the greenhouse and generated seed families for this experiment by haphazardly crossing individuals from the same population. We selected seed families to maximize the number of field-derived individuals represented. Thus, we used seeds from \Sexpr{nlevels(master$Family)} greenhouse-derived seed families, \Sexpr{min(famTab)}--\Sexpr{max(famTab)} (mean = \Sexpr{round(mean(famTab), 1)}, median = \Sexpr{median(famTab)}) families per population.

@@

# Rug indicating measurment timing
rect(6, -2, 17, 0, border = "white", col = "grey")
with(exptlDes, text(11.5, -1, labels = "Germ", col = "black", font = 3, 
  cex = 0.9))

with(exptlDes, rect(min(which(LERmeas)), -2, max(which(LERmeas)), 0, 
	border = "white", col = "grey"))
with(exptlDes, text(min((which(LERmeas)) + max(which(LERmeas))) / 2, -1, 
	labels = "LER", col = "black", font = 3, cex = 0.9))

with(exptlDes, rect(min(which(SERmeas)), 0, max(which(SERmeas)), 2,
	border = "white", col = "grey"))
with(exptlDes, text(min((which(SERmeas)) + max(which(SERmeas))) / 2 - 5, 1, 
	labels = "SER", col = "black", font = 3, cex = 0.9))

with(exptlDes, rect(min(which(Harvest)), -2, max(which(Harvest)), 0, 
	border = "white", col = "grey"))
with(exptlDes, text(min((which(Harvest)) + max(which(Harvest))) / 2, -1, 
	labels = "Harvest", col = "black", font = 3, cex = 0.9))

with(exptlDes, rect(min(which(PhotoMeas)), 2, max(which(PhotoMeas)), 4, 
	border = "white", col = "grey"))
with(exptlDes, text(min((which(PhotoMeas)) + max(which(PhotoMeas))) / 2, 5, 
	labels = "Photo", col = "black", font = 3, cex = 0.9))

@@

  # Confidence intervals on effect size (mean squared error)
  fn <- function(merMod) {
    x <- anova(merMod)
    ret <- x[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Mean Sq"]
    ret <- c(ret, x["PopID", "Mean Sq"] - x["PopID:TempTrt", "Mean Sq"],
             x["PopID", "Mean Sq"] - x["PopID:WaterTrt", "Mean Sq"])
    ret
  }

  ciMerModLER <- confint(mm, method = "boot", FUN = fn, nsim = 1e3, seed = 50717035,
                         parallel = "multicore", ncpus = 2L, level = 0.95)
  ciMerModLER <- as.data.frame(ciMerModLER)
  fit <- anova(mm)
  ciMerModLER$estimate <- c(fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Mean Sq"],
                            fit["PopID", "Mean Sq"] - fit["PopID:TempTrt", "Mean Sq"],
                            fit["PopID", "Mean Sq"] - fit["PopID:WaterTrt", "Mean Sq"])
  rownames(ciMerModLER) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
                             "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt")
  rm(fit)

@@

  # Confidence intervals on effect size (mean squared error)
  ciMerModSER <- confint(mm, method = "boot", FUN = fn, nsim = 1e3, seed = 46887099,
                         parallel = "multicore", ncpus = 2L)
  ciMerModSER <- as.data.frame(ciMerModSER)
  fit <- anova(mm)
  ciMerModSER$estimate <- c(fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Mean Sq"],
                            fit["PopID", "Mean Sq"] - fit["PopID:TempTrt", "Mean Sq"],
                            fit["PopID", "Mean Sq"] - fit["PopID:WaterTrt", "Mean Sq"])
  rownames(ciMerModSER) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
                             "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt")
  rm(fit)

@@

\paragraph{Photosynthesis}
During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate on a subset of \Sexpr{length(which(!is.na(master$Photo)))} plants evenly spread between treatments and families within populations. The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6-cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska). We made all measurements at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$ of photosynthetically active radiation), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity. All measurements were taken between 9:00 AM and 5:00 PM (3 hours after lights turned on and 5 hours before lights turned off). During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to the average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$) so that all plants within a temperature level could be measured under the same conditions. We were photosynthesis after dry down had progressed to assess differences in photosynthetic responses to drought.

@@

  # Confidence intervals on effect size (mean squared error)
  ciMerModPhoto <- confint(mm, method = "boot", FUN = fn, nsim = 1e3, seed = 12657909,
                           parallel = "multicore", ncpus = 2L)
  ciMerModPhoto <- as.data.frame(ciMerModPhoto)
  fit <- anova(mm)
  ciMerModPhoto$estimate <- c(fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Mean Sq"],
                              fit["PopID", "Mean Sq"] - fit["PopID:TempTrt", "Mean Sq"],
                              fit["PopID", "Mean Sq"] - fit["PopID:WaterTrt", "Mean Sq"])
  rownames(ciMerModPhoto) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
                               "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt")
  rm(fit)

@@

  fn <- function(merMod) {
    x <- Anova(merMod, type = 2)
    ret <- x[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Chisq"]
    ret <- c(ret, x["PopID", "Chisq"] - x["PopID:TempTrt", "Chisq"],
             x["PopID", "Chisq"] - x["PopID:WaterTrt", "Chisq"])
    ret
  }

  system.time(
  ciMerModMort <- confint(fitMort, method = "boot", FUN = fn, nsim = 1e1, seed = 925200709,
                          parallel = "multicore", ncpus = 2L)
)
  # Figure of reaction norms
  d <- subset(master, !is.na(master$DiedFromStress))
  tab <- tapply(d$DiedFromStress, list(d$PopID, d$WaterTrt, d$TempTrt), 
                function(X) 1 - length(which(X)) / length(X))
  pdf(paste(pathFig, "/FigureS_MortRxnNorm.pdf", sep = ""), 4, 4, useDingbats = F)
  par(mai = c(0.5, 1, 0.5, 0.25))
  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, 
       xlim = c(1, 4), ylim = c(0, 1))
  abline(v = 2.5)
  mtext(side = 3, c("Cool", "Hot"), at = c(1.625, 3.375))
  mtext(side = 1, c("Wet", "Dry", "Wet", "Dry"), at = c(1, 2.25, 2.75, 4))
  mtext(side = 1, "Water Treatment", font = 2, cex = 1.25, line = 1.5)
  mtext(side = 3, "Temperature Treatment", font = 2, cex = 1.25, line = 1.5)
  axis(2, lwd = 0, lwd.ticks = 1, las = 1)
  title(ylab = "Survivorship", cex.lab = 1.5)

@@

<<Effect size confidence intervals, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=

  pathFig <- "/Users/cdmuir/Google Drive/CardAdapt/ms/Figures"

pdf(paste0(pathFig, "/FigureS_EffectSizes.pdf"), 12, 4, useDingbats = F)
par(mai = c(1, 1, 0.5, 0.1), mfrow = c(1, 3))

for (mod in list(ciMerModLER, ciMerModSER, ciMerModPhoto)) {

#  assign("mod", mod)

  plot(0, 0, type = "n", axes = FALSE, frame.plot = TRUE,
       ylim = range(mod), xlim = c(1, 5),
       xlab = "", ylab = "Effect size (Mean squared error)", cex.lab = 1.5,
       main = )

  abline(h = 0, lty = 2, col = "grey50")
  for (i in 1:5) {
    lines(rep(i, 2), c(mod[i, "2.5 %"], mod[i, "97.5 %"]))
    points(i, mod[i, "estimate"], pch = 19)
  }

  axis(2, las = 1, lwd = 0, lwd.ticks = 1)
  clip(-100, 100, -100, 100)
  text(x = 1:5, y = grconvertY(-0.025, "npc", "user"), srt = 45, pos = 2, 
       labels = c("Pop", "Pop x Temp", "Pop x Water", "Pop - Pop x Temp",
                  "Pop - Pop x Water"))

}

\subsection*{Genetic variation in trait means and plasticity}

For all traits (Table~\ref{table:Table_traits}) except germination (see above), we tested for Population, Treatment (Temperature, Water, and Temperature $\times$ Water), and Population $\times$ Treatment interactions (Population $\times$ Temperature, Population $\times$ Water, and Population $\times$ Temperature $\times$ Water). We interpreted significant Population effects to indicate genetic variation in trait means and Population $\times$ Treatment interactions to indicate genetic variation in plasticity. As mentioned above, we used survival and GLMM models for germination rate and mortality, respectively. For all other traits, we used mixed model ANOVAs with Family and Block included as random factors. We fit models using restricted maximum likelihood in lmer, a function in the R package \pkg{lme4} \citep{Bates_etal_2015}. We determined significant fixed effect terms using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-32 \citep{Kuznetsova_etal_2016}. We used Satterthwaite's approximation to calculate denominator degrees of freedom for $F$-tests. We also included days to germination as a covariate in growth analyses. To ensure that Population and Treatment effects were specific to a particular growth phase, we included germination day as a covariate in leaf expansion and stem elongation analyses.

Failure to detect a significant effect could be the result of Type-2 error, so we complemented the above analysis by comparing effect sizes in the full model. The full model contains all main effects, two-way interactions (Population $\times$ Temperature, Population $\times$ Water), a three-way interaction (Population $\times$ Temperature $\times$ Water), and random effects. We compared the effect sizes (mean squared error for LMM, ? for GLMM) of Population to Population $\times$ Treatment in 1000 parametric bootstrap samples simulated from fitted models. MARKER

@@

We selected Climate-Latitude and Climate-TraitPC1 variables independently using Variable Selection Using Random Forest (VSURF) algorithm in the R package \pkg{VSURF} version 1.0.3 \citep{Genuer_etal_2016}. Random Forest regression is useful for cases like ours when the number of potential predictors is similar to or greater than the number of observations (`high $p$, low $n$' problem). VSURF is a multistep algorithm that progressively retains or eliminates variables based on their importance over regression trees in the forest. Variable importance is defined as the average amount a climate variable reduces mean-squared error in the predicted response (TraitPC1 or Latitude), compared to a randomly permuted dataset, across all trees in the random forest (see \citeauthor{Genuer_etal_2015} [\citeyear{Genuer_etal_2015}] for further detail). Hence, VSURF automatically eliminates unimportant and redundant variables based on the data without having to arbitrarily choose among colinear climate variables before the analysis. We kept only variables selected for prediction, the most stringent criterion. We visually depict how we selected climatic variables in Fig~\ref{fig:Fig_VarSelect}.

<<Fig_VarSelect, eval=FALSE, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=

@@

In contrast to the genetic variation in trait means described above, we found little evidence of G$\times$E in \textit{E. cardinalis}. There were only two statistically significant Population $\times$ Treatment interactions (Table~\ref{table:Table_AnovaSummary}, Fig.~\ref{fig:FigS_RxnNorms}), but these were not strong compared to Population and Temperature effects. Otherwise, populations responded similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}). Complete ANOVA tables are available in the Supporting Information (\Cref{table:TableS_fitLER,table:TableS_fitSER,table:TableS_fitPhoto,table:TableS_fitMort})

@@

Interannual variation in climate averaged over each populations's climatic neighborhood correlated most strongly with trait variation and latitude of \textit{E. cardinalis} occurrences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}). All \Sexpr{nImpLat} Climate-Latitude and \Sexpr{nImpPC1} Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}). In fact, neighborhood climate almost always correlated better with TraitPC1 and Latitude than local climate (Fig.~\ref{fig:Fig_LocalvNeighCorr}). On average, neighborhood Climate-TraitPC1 correlation coefficients were \Sexpr{-round(tt1$estimate, 2)} higher than correlations with local-scale climate variables (paired \textit{t}-test, $t = \Sexpr{-round(tt1$statistic, 2)}$, d.f. = \Sexpr{round(Meff1 / 2 - 1, 1)}, $P = \Sexpr{round(pt1 * 10 ^ -floor(log10(pt1)), 2)} \times 10 ^ {\Sexpr{floor(log10(pt1))}}$). Likewise, neighborhood Climate-Latitude correlation coefficients were \Sexpr{-round(tt2$estimate, 2)} higher than those for local-scale climate (paired \textit{t}-test, $t = \Sexpr{-round(tt2$statistic, 2)}$, d.f. = \Sexpr{round(Meff2 / 2 - 1, 1)}, $P = \Sexpr{round(pt2 * 10 ^ -floor(log10(pt2)), 2)} \times 10 ^ {\Sexpr{floor(log10(pt2))}}$). 
Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates to explain the latitudinal cline in \textit{E. cardinalis} (see Table~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations). Note that the coefficient of variation of a climatic factor is subscripted with $\sigma$ whereas the mean is subscripted with $\mu$. More specifically, greater winter precipitation variability and lower winter temperature variability are associated with Southern latitudes and higher TraitPC1 values (Fig.~\ref{fig:Fig_ClimPC1}A,B). Neighborhood interannual variation in winter precipitation (bio16$_\sigma$) was the most important Climate-Latitude variable (Fig.~\ref{fig:Fig_ClimVarImp}A). However, neighborhood bio16$_\sigma$ did not overlap with Climate-TraitPC1 variables (Fig.~\ref{fig:Fig_ClimVarImp}B). We nevertheless consider it a plausible candidate for two reasons. First, neighborhood bio16$_\sigma$ correlated strongly with TraitPC1 (Fig.~\ref{fig:Fig_ClimPC1}A). Second, one of the most important Climate-TraitPC1 variables (neighborhood bio15$_\sigma$; Fig.~\ref{fig:Fig_ClimPC1}B,C) is very similar to bio16$_\sigma$. In Mediterranean climates like California, most precipitation occurs in the wettest quarter (winter), so years with low winter precipitation also have low precipitation seasonality. Hence, highly variable year-to-year winter precipitation at lower latitude (Fig.~\ref{fig:Fig_ClimPC1}D) is closely associated with large swings in precipitation seasonality (Fig.~\ref{fig:Fig_ClimPC1}C).
